---
title: "R Notebook"
output: html_notebook
---



```{r}
library(data.table)
library(magrittr)
library(tidyr)
```
### Section 01

 For questions 1,2 and 3, look at the pdf with the answers 

 4. The example_product_data.csv file describes the number of times a person bought product “a” and “b”. Load the file into a data.table. 

```{r}
product_dt <- fread("extdata/example_product_data.csv")
product_dt
```

 5. Transform product_dt into a long fromat using data.table commands.
```{r}
long_dt <- melt(product_dt, id.vars = "name", measure.vars =c("producta", "productb"), variable.name = "product", value.name = "num_prod" )
long_dt
```

6. Transform the table from the long format back into a wide format. Check that it is equal to the original data.table.
```{r}
wide_dt<- dcast(long_dt, ... ~ product,value.var ="num_prod" )
wide_dt
```

### Section 02

1. Read in the weather dataset weather.txt as a data.table.
```{r}
weather_dt<- fread("extdata/weather.txt")
head(weather_dt)
```

2. Why is this dataset messy?

- Column headers are values, not variable names (d1,d2...) so we need to melt.
- Multiple variables are stored in one column (element), so we need to dcast.
- It would be better if year, month and day were in the same column called "date". We need to unite

3. How would a tidy version of it look like? Do not give the code, only describe how the tidy table would look like.

It would be something like this:
      

```{r}
DT = data.table(
  id = c("MX000017004","MX000017004"),
  date = c("01.12.2010","02.02.2010"),
  tmin = c(138,144),
  tmax = c(299,273)
) 

DT
```

4. Create a tidy version of the weather dataset.
<br>
id.vars have all the variables that do not change. The rest are considered to be melted.
```{r}
melted_weather <- melt(weather_dt, id.vars = c("id","year","month","element"), variable.name = "day", value.name = "temperature")
melted_weather <- melted_weather[!(is.na(temperature))]
head(melted_weather)

weather_dcast<- dcast(melted_weather, ... ~ element,  value.var = "temperature")
head(weather_dcast)

```

```{r}

weather_dcast[,day := as.integer(gsub('d','',day))] 
head(weather_dcast)

weather_dcast[,date := paste(day, month, year,sep = "-")]
weather_dcast[,c("year","month","day"):= NULL]
head(weather_dcast)

```
```{r}
setnames(weather_dcast,"TMAX","tmax")
setnames(weather_dcast,"TMIN","tmin")
head(weather_dcast)
```


### Section 03 - Scattered data across many files

1. Create a list containing all file paths in the folder.

```{r}
files <- list.files("extdata/baby-names/", full.names = T)
head(fread(files[1]))

```

2. Name the list entries with the basename of the corresponding file path.
```{r}
names(files) <- basename(files)
```

3. Read in the data from all files into one table. Hint: when you read many files and gather them into one table, be sure to add a column that identifies each file. rbindlist()
```{r}
tables <- lapply(files, fread)
babies_dt <- rbindlist(tables, idcol = 'filename')
head(babies_dt)
```

4. Is the data tidy? If not, tidy it up.
```{r}
#We want two columns called sex and year
babies_dt <- separate(babies_dt, col = "filename", into=c("year","sex"), extra="drop")
head(babies_dt)
```

### Section 04 - Merge Warm Up

Prepare two tables by running the following code:
```{r}
mtcars_dt <- as.data.table(mtcars)
mtcars_dt[, carname := rownames(mtcars)]
dt1 <- mtcars_dt[5:25,.(carname, mpg, cyl)]
dt2 <- mtcars_dt[1:10, .(carname, gear)]
head(dt1)
head(dt2)
```

1. How long is the inner merge of dt1 and dt2?
```{r}
inner_dt <- merge(dt1, dt2, by='carname')
inner_dt
```
2. How long is the left merge of dt1 and dt2?
```{r}
left_dt <- merge(dt1, dt2, by='carname',all.x = T)
left_dt
```

3. How long is the outter merge of dt1 and dt2?
```{r}
outter_dt <- merge(dt1, dt2, by='carname',all = T)
outter_dt
```
### Section 05 - Small case study

Read the two files in the folder eqt1
```{r}
gt <- fread("extdata/eqtl/genotype.txt")
dim(gt)
```

```{r}
head(gt[,1:5])
```
```{r}
growth <- fread('extdata/eqtl/growth.txt')
head(growth)
```

 2. Come up with a strategy, how you can transform the two tables shown above into the single table shown below.
<br>
A table with column names: strain, media, growth_rate, marker and gt.
```{r}
gt <- melt(gt, id.vars = "strain", variable.name = "marker", value.name = "gt")
head(gt)
```


```{r}
growth <- melt(growth, id.vars = "strain", variable.name = "media", value.name = "growth_rate")
head(growth)
```
```{r}
final_dt <- merge(growth, gt, by='strain',allow.cartesian = T)
head(final_dt)
summary(final_dt)
```

```{r}
library(ggplot2)
ggplot(final_dt[marker %in% c('mrk_5211', 'mrk_1653')], aes(marker, growth_rate, color=gt)) + geom_boxplot() + facet_wrap(~media)
```

