---
title: "R Notebook"
output: html_notebook
---

Since we are using data.tables the $ sign should not be used! We should use the DT env [] as much as you can.

### Section 00 - Getting ready

```{r}
library(data.table)
library(magrittr)
```


### Section 01 - Reading and cleaning up data

1. Load the three given datasets as data.tables and name them as users_dt, books_dt and ratings_dt accordingly. Hint: fread()
```{r}
users_dt<- fread("./BX-Users.csv")
ratings_dt<-fread("./BX-Book-Ratings.csv")
books_dt <- fread("./BX-Books.csv")
```

2. Check the classes of users_dt, ratings_dt and books_dt. Confirm that these are indeed a data.table.
```{r}
class(users_dt)
class(ratings_dt)
class(books_dt)
```


3. Check the column names and classes of the users_dt data table and change the type of the Age column in users_dt to numeric.
```{r}
colnames(users_dt)
sapply(users_dt, class)
users_dt[,Age:= as.numeric(Age)]
sapply(users_dt, class)
```

4. Produce a summary of the variables in books_dt.
```{r}
summary(books_dt)

```

5. Return the first 5 and last 5 observations of the table ratings_dt.
```{r}
head(ratings_dt)
tail(ratings_dt)

```

6. Replace all the - in column names by underscores _ in all three data tables. For example, Book-Title should be renamed to Book_Title. Hint: You can use the function gsub() that replaces pattern in a character string by a defined replacement. For example, for replacing R by DataViz in the following sentence
s we use:
```{r}
s <- 'R is fun'
gsub('R', 'DataViz', s)

```

```{r}
colnames(books_dt)
colnames(books_dt) <- gsub('-', '_', colnames(books_dt))
colnames(books_dt)

colnames(ratings_dt) <- gsub('-', '_', colnames(ratings_dt))
colnames(users_dt) <- gsub('-', '_', colnames(users_dt))
```

7. Delete the columns Image-URL-S, Image-URL-M and Image-URL-L in the table books_dt.
```{r}
books_dt[, c("Image_URL_S", "Image_URL_M", "Image_URL_L"):= NULL]

```

8. What is the first year of publication? What is the last one?
```{r}
books_dt[,min(Year_Of_Publication)]
books_dt[,max(Year_Of_Publication)]

```

9. Remove all the books published before 1900 and later than 2019 from books_dt.
```{r}
dim(books_dt)
books_dt <- books_dt[Year_Of_Publication >= 1900 & Year_Of_Publication <=2019]
dim(books_dt)
```

### Section 02 - Data Exploration

1. How many different authors are included in the table books_dt?
```{r}
books_dt[, unique(Book_Author)] %>% length()

#Other option: books_dt[, uniqueN(Book_Author)]
```

2. How many different authors are included for each year of publication between 2000 and 2010 in books_dt?
```{r}
books_dt[Year_Of_Publication>=2000 & Year_Of_Publication<=2010,uniqueN(Book_Author),by=Year_Of_Publication][order(-Year_Of_Publication)]
#uNIQUEn
```


3. In how many observations is the age information missing in the users table users_dt?.
```{r}
#users_dt[,is.na(Age)] %>% length()

users_dt[is.na(Age), .N]
```

4. Have a look at all locations from teenager users the table users_dt.
```{r}
users_dt[Age>=13 & Age<=19, unique(Location)] %>% head(n=10)
#Use head because it is long

```

5. What is the maximum rating value in the ratings table?.
```{r}
ratings_dt[,max(Book_Rating)]
ratings_dt[,min(Book_Rating)]
```

6. What is the most common rating value larger than 0?
```{r}
ratings_dt[Book_Rating>0,.N,by=Book_Rating][N==max(N)]

#.N always goes in second position
```

```{r}

```

7. Which are the book identifiers (ISBN) with the highest ratings?
```{r}
ratings_dt[Book_Rating == max(Book_Rating), ISBN] %>% head()
#We use head because it is too long again
```

8. Sort the ratings table according to the rating value of each book in descending order. Hint: order()
```{r}
ratings_dt <- ratings_dt[order(-Book_Rating)]
head(ratings_dt)
```

9. Create a new column Country in the table users_dt for the name of the country of each user. For
instance, from the location cologne, nrw, germany, we can assume the user comes from Germany. Hint:
tstrsplit()
```{r}
users_dt[, Country:= tstrsplit(Location, ",")[[3]]]
?tstrsplit
```

10. How many different countries are contained in the table users_dt?
```{r}
users_dt[,uniqueN(Country)]
#The number is too big, it is not our fault but the data table fault. We would have to clean it

```

11. What is the average age of the users in users_dt? What is the average age for users in NYC, Stockton
and Moscow? Hint: use by:= and i for row filtering
```{r}

users_dt[, City:= tstrsplit(Location, ",")[[1]]]
#users_dt[, City:= tstrsplit(Location, ",", keep=1)]

users_dt[City %in% c("nyc","stockon", "moscow"), mean(Age, na.rm = TRUE), by=City]
```

### Section 03 - Manipulating data tables

1. Add a new column called High_Rating to the data table ratings_dt. The column has an integer 1 for all observations with a rating value higher than 7
```{r}

ratings_dt[Book_Rating>7, High_Rating := 1]
ratings_dt[Book_Rating<=7, High_Rating := 0]
ratings_dt %>% head()

#Other way
#ratings_dt[,High_Rating := ifelse (Book_Rating>7, 1, 0)]

```

2. How many observations are considered to be a high ranking? What is the proportion of high ranked observations among all observations?
```{r}
ratings_dt[,.N,by=High_Rating]
ratings_dt[High_Rating==1,.N]/ratings_dt[,.N]
#Other way
#ratings_dt[,sum(High_Rating)/.N]
```

3. Set the book identifier the key of the data table books_dt. What happened to the order of the data table? Hint: setkey()
```{r}
setkey(books_dt,"ISBN")
#It changes the order of the rows
```

4. Which users did not give any rating to any book? Filter these users out from users_dt. Hint: There’s no need to merge users_dt with ratings_dt, we are simply interested in the users that are not in ratings_dt.
```{r}

users_who_rated <- ratings_dt[,User_ID]
users_dt[!User_ID %in% users_who_rated] %>% head()
#users_not_rate <- users_dt[!User_ID %in% users_who_rated]

```

5. What is the most common age of users who rated at least one book?
```{r}
users_dt[User_ID %in% users_who_rated & !is.na(Age), .N, by=Age][N == max(N)]
```

6. On average, how many books did a user rate?
```{r}
ratings_dt[,.N, by=User_ID][,mean(N, na.rm=T)]
```



7. What is the title of the first published book with the highest ranking?
```{r}
ratings_dt[order(Year_Of_Publication, -Book_Rating)] %>% head(1)
```

8. In which year was a book with the largest number of ratings last published?
```{r}
ratings_dt[,Rating_Count := .N, by=ISBN]
ratings_dt[Rating_Count==max(Rating_Count), max(Year_Of_Publication)] 
```

9. Add to the table ratings_dt the highest ranking that each book received as a new column called Max_Book_Ranking.
```{r}
ratings_dt[,Max_Rating := max(Book_Rating), by=ISBN]
```

10. Subset the merged ratings table to contain only books written by the following authors:
```{r}
authors <- c("Agatha Christie", "William Shakespeare", "Stephen King", "Ann M. Martin", "Carolyn Keene", "Francine Pascal",
"Isaac Asimov", "Nora Roberts", "Barbara Cartland", "Charles Dickens")

ratings_dt[Book_Author %in% authors, .(mean(Book_Rating, na.rm=T), max(Book_Rating)), by=Book_Author]

```

